---
title: "TempEst 2 Demo"
author: "Daniel Philippus"
date: "`r format(Sys.Date(), format='%b %d, %Y)')`"
output: pdf_document
---

# TempEst 2 User Guide

This document provides a comprehensive guide for using the TempEst 2 (SCHEMA)
stream water temperature estimation model. TempEst 2 estimates stream water
temperature at any point using satellite remote sensing data.

**Author's Note (Justin):** This codebase was developed during my time at CIROH.
Below you'll find detailed instructions on how to use each component.

---

## Table of Contents
1. [Prerequisites](#prerequisites)
2. [Python Data Retrieval Scripts](#python-data-retrieval-scripts)
3. [Data Format Requirements](#data-format-requirements)
4. [Running the Model](#running-the-model)
5. [Validation and Visualization](#validation-and-visualization)

---

## Prerequisites

### R Dependencies
Install the following R packages before running this notebook:

```r
install.packages(c("tidyverse", "fields", "geosphere", "lubridate", "zoo",
                   "patchwork", "dataRetrieval", "sf", "gstat", "sp",
                   "Metrics", "purrr", "effsize"))
```

### Python Dependencies
For data retrieval from Google Earth Engine:

```bash
pip install earthengine-api
```

### Google Earth Engine Setup

**Option A: Interactive Authentication (Recommended for new users)**

For most users, you can authenticate interactively without needing a service account:

```python
import ee
ee.Authenticate()  # Opens browser for Google login
ee.Initialize()    # Initializes with your credentials
```

This is the simplest method - just run `ee.Authenticate()` once, and it will open
your browser to log in with your Google account. After that, `ee.Initialize()`
will work automatically.

**Option B: Service Account Credentials (For automated/server use)**

If you need to run scripts on a server or want persistent credentials, use a
service account:

1. Go to Google Cloud Console: https://console.cloud.google.com/
2. Create a new project (or select existing)
3. Enable the Earth Engine API for your project
4. Go to "IAM & Admin" > "Service Accounts"
5. Create a new service account
6. Create a JSON key for the service account (downloads automatically)
7. Register the service account with Earth Engine at:
   https://signup.earthengine.google.com/#!/service_accounts

Then update the Python scripts:
```python
service_account = 'your-account@your-project.iam.gserviceaccount.com'
credentials = ee.ServiceAccountCredentials(service_account, '/path/to/your-key.json')
ee.Initialize(credentials)
```

**Note:** The current scripts use hardcoded paths to a service account JSON file
(e.g., `/Users/justin/Downloads/hogue-428318-8f3882796b90.json`). You will need
to either:
- Replace these paths with your own service account JSON path, OR
- Comment out the service account lines and use `ee.Authenticate()` + `ee.Initialize()` instead

---

## Python Data Retrieval Scripts

### Overview of Scripts

There are **three main Python scripts** for retrieving environmental data:

| Script | Purpose | Data Source | Output |
|--------|---------|-------------|--------|
| `eeretrieval.py` | Get LST, humidity, elevation, land cover | MODIS, NLDAS, ESA WorldCover | TempEst predictor data |
| `precipitation.py` | Get daily precipitation | NASA NLDAS | precipitation CSV |
| `weather.py` | Get cloud cover | NOAA NCEP DOE R2 | cloud cover CSV |

---

### 1. eeretrieval.py - Main Earth Engine Data Retrieval

**Purpose:** Retrieves all predictor variables needed for TempEst 2 predictions.

**Variables Retrieved:**
- Land Surface Temperature (LST) from MODIS
- Specific Humidity from NLDAS
- Elevation from MERIT DEM
- Land cover abundances (water, shrubland, grassland, barren) from ESA WorldCover

**How to Use:**

```python
# 1. First, define your points of interest in newdatapts.py
#    Format: [[[longitude, latitude], "point_id"], ...]
#    Example:
potomacdatapts = [
    [[-78.56894, 39.52073], '1'],
    [[-78.51559, 39.52358], '2'],
    # ... more points
]

# 2. Update eeretrieval.py with your credentials:
service_account = 'your-service-account@your-project.iam.gserviceaccount.com'
credentials = ee.ServiceAccountCredentials(service_account, 'path/to/credentials.json')

# 3. Set your date range:
times = [(str(x), str(y+1),
          str(dt.date(x, 1, 1) + dt.timedelta(y)),
          str(dt.date(x, 1, 1) + dt.timedelta(y+1)))
         for x in range(2020, 2024)  # Years to retrieve
         for y in range(365)]         # Days of year

# 4. Run the script:
python eeretrieval.py
```

**Output:** CSV files saved to Google Drive folder (e.g., "MississippiData/AllData2020_1.csv")

**Important Notes:**
- Maximum 3000 GEE jobs can run at once
- Use `wait` parameter to space out job submissions (20 seconds works for ~1300 points)
- Data is saved to Google Drive, download and combine CSVs manually

---

### 2. precipitation.py - Precipitation Data

**Purpose:** Retrieves daily total precipitation from NASA NLDAS.

**How to Use:**

```python
# 1. Import your data points from newdatapts.py
from newdatapts import potomacdatapts

# 2. Update credentials (same as eeretrieval.py)

# 3. Set date range:
start_date = dt.date(2020, 1, 1)
end_date = dt.date(2023, 12, 31)

# 4. Set output file name:
output_file = "precipitation_potomac.csv"

# 5. Run the script:
python precipitation.py
```

**Output CSV Format:**
| point_id | lon | lat | date | precipitation |
|----------|-----|-----|------|---------------|
| Point1 | -78.56 | 39.52 | 2020-01-01 | 0.5 |

---

### 3. weather.py - Cloud Cover Data

**Purpose:** Retrieves daily mean cloud cover from NOAA NCEP DOE Reanalysis 2.

**How to Use:**

```python
# 1. Import your data points from newdatapts.py
from newdatapts import potomacdatapts, sturgeon_pts

# 2. Update credentials (same as eeretrieval.py)

# 3. Set time range in the `times` variable:
times = [(str(year), str(day + 1),
          str(dt.date(year, 1, 1) + dt.timedelta(days=day)),
          str(dt.date(year, 1, 1) + dt.timedelta(days=day + 1)))
         for year in range(2020, 2023)
         for day in range(365)]

# 4. Set output file:
output_file = "cloud_cover_potomacriver.csv"

# 5. Run the script:
python weather.py
```

**Output CSV Format:**
| point_id | lon | lat | date | mean_cloud_cover |
|----------|-----|-----|------|------------------|
| Point 1 | -78.56 | 39.52 | 2020-01-01 | 45.2 |

---

## Defining Points of Interest (newdatapts.py)

Create your river points in `newdatapts.py`. Each river needs a list of
coordinate points along the stream reach.

**Format for eeretrieval.py and precipitation.py:**
```python
river_pts = [
    [[-longitude, latitude], "PointID"],
    [[-78.56894, 39.52073], '1'],
    [[-78.51559, 39.52358], '2'],
    # ... continue for all points
]
```

**Format for weather.py (note the nested list):**
```python
river_pts = [
    [[-88.4918, 47.0322], "Point 1"],
    [[-88.4983, 47.0291], "Point 2"],
    # ...
]
```

**Tips for defining points:**
- Space points evenly along the river reach (e.g., every ~1km)
- Include 50-100 points per river for good spatial coverage
- Use Google Earth or QGIS to digitize points along the stream centerline

---

## Data Format Requirements for TempEst 2

After retrieving data from Google Earth Engine, combine all CSVs and ensure
your data frame has these columns:

| Column | Type | Description |
|--------|------|-------------|
| id | character | Point identifier |
| lon | numeric | Longitude (decimal degrees East) |
| lat | numeric | Latitude (decimal degrees North) |
| date | Date | Date of observation |
| day | integer | Julian day (1-366) |
| elevation | numeric | Mean elevation in 500m radius (meters) |
| lst | numeric | Land surface temperature (Celsius) |
| humidity | numeric | Specific humidity (kg/kg) |
| water | numeric | Water land cover abundance (0-1) |
| shrubland | numeric | Shrubland abundance (0-1) |
| grassland | numeric | Grassland abundance (0-1) |
| barren | numeric | Barren land abundance (0-1) |

**Example R code to prepare data:**
```r
data <- map_dfr(
  list.files("data/YourRiverData", full.names = TRUE),
  \(fn) read_csv(fn, col_types="cDddddddddd") %>%
    mutate(day = as.integer(format(date, "%j"))) %>%
    filter(lst > -273)  # Filter invalid LST values
)
```

---

## Complete Execution Pipeline

Follow these phases IN ORDER. Each phase depends on the previous one.

```
┌─────────────────────────────────────────────────────────────────────────┐
│                        EXECUTION PIPELINE                                │
├─────────────────────────────────────────────────────────────────────────┤
│                                                                          │
│  PHASE 0: Data Collection (Python - Run BEFORE opening this notebook)   │
│  ├── 0.1 Define points in newdatapts.py                                  │
│  ├── 0.2 Run eeretrieval.py → Downloads to Google Drive                  │
│  ├── 0.3 Run precipitation.py → Creates precipitation CSV                │
│  ├── 0.4 Run weather.py → Creates cloud cover CSV                        │
│  └── 0.5 Download CSVs from Google Drive to data/ folder                 │
│                                                                          │
│  PHASE 1: Setup (R - Lines ~277-294)                                     │
│  └── Load all R libraries                                                │
│                                                                          │
│  PHASE 2: Load Earth Engine Data (R - Lines ~312-336)                    │
│  └── Read predictor CSVs from data/ folders                              │
│                                                                          │
│  PHASE 3: Load Cloud Cover & Gage Data (R - Lines ~339-364)              │
│  ├── Read cloud cover CSVs                                               │
│  └── Read USGS gage data (or download via dataRetrieval)                 │
│                                                                          │
│  PHASE 4: Run TempEst Model (R - Lines ~824-835)                         │
│  ├── load("model.rda")                                                   │
│  └── predictions <- model(data)                                          │
│                                                                          │
│  PHASE 5: Merge All Data (R - Lines ~836-873)                            │
│  └── Join predictions + cloud cover + precipitation                      │
│                                                                          │
│  PHASE 6: Add Weather Conditions (R - Lines ~636-691)                    │
│  └── Categorize: "Cloudy", "Raining and Cloudy", "No Clouds or Rain"     │
│                                                                          │
│  PHASE 7: Interpolation (R - Lines ~875-990)                             │
│  └── Fill gaps in time series using linear interpolation                 │
│                                                                          │
│  PHASE 8: Validation & Visualization (R - Lines ~992-2068)               │
│  ├── Compare gage vs model temperatures                                  │
│  ├── Statistical analysis (RMSE, MAE, Cohen's d)                         │
│  └── Generate plots                                                      │
│                                                                          │
│  PHASE 9: Training New Model (Optional - Lines ~2323-2547)               │
│  └── model <- full.schema()(training_data)                               │
│                                                                          │
│  PHASE 10: Machine Learning (Optional - Lines ~2734-2777)                │
│  └── Train RNN using keras                                               │
│                                                                          │
└─────────────────────────────────────────────────────────────────────────┘
```

---

## Detailed Step-by-Step Instructions

### PHASE 0: Data Collection (Python)

**Run these BEFORE opening demo.Rmd in RStudio.**

#### Step 0.1: Define Your River Points
Edit `newdatapts.py` to add your river coordinates:
```python
my_river_pts = [
    [[-longitude, latitude], "PointID"],
    # Add 50-100 points along your river
]
```

#### Step 0.2: Retrieve Main Predictor Data
```bash
# Edit eeretrieval.py:
# - Update service_account and credentials path
# - Set date range in 'times' variable
# - Import your points from newdatapts.py
python eeretrieval.py
```
This exports CSVs to Google Drive. **Wait for GEE jobs to complete**, then
download all CSVs and place in `data/YourRiverData/` folder.

#### Step 0.3: Retrieve Precipitation Data
```bash
# Edit precipitation.py:
# - Update the import to use your points
# - Set start_date and end_date
# - Set output_file name
python precipitation.py
```
Creates: `precipitation_yourriver.csv`

#### Step 0.4: Retrieve Cloud Cover Data
```bash
# Edit weather.py:
# - Update the import to use your points
# - Set years in 'times' variable
# - Set output_file name
python weather.py
```
Creates: `cloud_cover_yourriver.csv`

---

### PHASE 1-8: R Analysis (In This Notebook)

Run the code chunks in order. Key sections:

| Phase | What It Does | Key Variables Created |
|-------|--------------|----------------------|
| 1 | Load libraries | - |
| 2 | Load GEE predictor data | `dataSR`, `data2`, `data3`, `data4` |
| 3 | Load cloud/gage data | `cloud_dataSR`, `gage_dataSR` |
| 4 | Run TempEst model | `predSR`, `pred2`, `pred3`, `pred4` |
| 5 | Merge predictions + cloud | `merged_dataSR`, `merged_data2`, etc. |
| 6 | Add weather categories | `merged_data_weatherSR` |
| 7 | Interpolate gaps | `df_interp`, `df_interp_gage` |
| 8 | Validation plots | Various ggplot outputs |

---

### Quick Start for New River

If you want to analyze a **new river**, here's the minimum you need:

1. **Python (before R):**
   ```bash
   # Add points to newdatapts.py, then:
   python eeretrieval.py      # Wait for GEE, download CSVs
   python precipitation.py
   python weather.py
   ```

2. **R (in this notebook):**
   ```r
   # Phase 1: Load libraries (run setup chunk)
   source("model.R")

   # Phase 2: Load your GEE data
   my_data <- map_dfr(
     list.files("data/MyRiverData", full.names = TRUE),
     \(fn) read_csv(fn) %>% mutate(day = as.integer(format(date, "%j")))
   )

   # Phase 3: Load cloud/gage data
   cloud_data <- read.csv("cloud_cover_myriver.csv")
   gage_data <- read_csv("MyRiverGageData.csv")

   # Phase 4: Run model
   load("model.rda")
   predictions <- model(my_data)

   # Phase 5: Merge
   merged <- predictions %>%
     left_join(cloud_data, by = c("id", "date"))
   ```

---

## Original Demo Content

This is a quick demo of how to use the TempEst 2 model code. Dependencies are
`tidyverse` and `fields`, both imported by `model.R`.

```{r setup, warning=FALSE, message=FALSE}
source("model.R")
library(tidyverse)
library(geosphere)
library(lubridate)
library(zoo)
library(patchwork)
library(dataRetrieval)
library(sf)
library(gstat)
library(sp)
library(dplyr)
library(tidyr)
library(Metrics)
library(purrr)
#statistical analysis
library(effsize)
```


# Earth Engine Data Retrieval

TempEst 2 comes with a Python script for automatically retrieving predictor
data from Google Earth Engine.  This requires that you have a Google Earth Engine
account and have local Earth Engine access set up for Python on your computer,
though it should be straightforward to modify for use in Google Collaboratory.

For that script, it is simply necessary to specify a date range of interest and
points of interest.  Examples are included for the format of the points of interest.
Then, the function `getAllTimeseries()` will retrieve a series of CSVs (one
for each timestep) in Google Earth Engine.  These CSVs should be joined together
into one data frame **before** making predictions, as TempEst 2 needs to analyze
long-term data to generate seasonality estimates.  Otherwise, the only modification
is to add a `day` column (e.g., `as.integer(format(date, "%j")))`).

```{r}
dataSR <- map_dfr(
  list.files("data/TempEstData", full.names = TRUE),
  \(fn) read_csv(fn, col_types="cDddddddddd") %>%
    mutate(day = as.integer(format(date, "%j"))) %>%
    filter(lst > -273)  # filters errors
)
data2 <- map_dfr(
  list.files("data/potomacriverdata", full.names = TRUE),
  \(fn) read_csv(fn, col_types="cDddddddddd") %>%
    mutate(day = as.integer(format(date, "%j"))) %>%
    filter(lst > -273)  # filters errors
)
data3 <- map_dfr(
  list.files("data/DRData", full.names = TRUE),
  \(fn) read_csv(fn, col_types="cDddddddddd") %>%
    mutate(day = as.integer(format(date, "%j"))) %>%
    filter(lst > -273)  # filters errors
)
data4 <- map_dfr(
  list.files("data/SRData", full.names = TRUE),
  \(fn) read_csv(fn, col_types="cDddddddddd") %>%
    mutate(day = as.integer(format(date, "%j"))) %>%
    filter(lst > -273)  # filters errors
)
```
#Gets cloud cover and gage data for River you provided 
```{r}
cloud_dataSR <- read.csv("cloud_cover_data.csv")
gage_dataSR <- read_csv("RoaringForkGageData.csv")



gage_dataSR <- gage_dataSR %>%
  mutate(date = as.character(date))
```

```{r}
cloud_data2 <- read.csv("cloud_cover_delawareriver.csv")
gage_data2 <- read_csv("DelawareDataGage.csv")
gage_data2 <- gage_data2 %>%
  mutate(date = as.character(date))


cloud_data3 <- read.csv("cloud_cover_sturgeonriver.csv")
gage_data3 <- read_csv("SturgeonDataGage.csv")
gage_data3 <- gage_data3 %>%
  mutate(date = as.character(date))

cloud_data4 <- read.csv("cloud_cover_potomacriver.csv")
gage_data4 <- read_csv("PotomacDataGage.csv")
gage_data4 <- gage_data4 %>%
  mutate(date = as.character(date))
```

```{r}
gage_data_date <- gage_data %>%
  left_join(cloud_data %>% select(date, mean_cloud_cover), by = "date") %>% 
  filter(date == ("2023-07-31"))
ggplot() +
  # Gage
  geom_point(data = gage_data_date, aes(x = lon, y = lat, color = temperature, size = mean_cloud_cover), shape = 17) +
  # TempEst
  geom_point(data = filtered_data_date, aes(x = lon.x, y = lat.x, color = temp.mod, size = mean_cloud_cover), shape = 16, alpha = 0.6) +
  scale_color_viridis_c(option = "plasma", name = "Temperature (C)") +
  scale_size_continuous(range = c(2, 10), name = "Cloud Cover") +
  theme_minimal() +
  labs(
    title = paste("Gage vs Model Temperature with Cloud Cover on", gage_data_date$date),
    x = "Longitude",
    y = "Latitude",
    color = "Temperature (°C)",
    size = "Mean Cloud Cover (%)"
  )
```
```{r}
gage_data <- gage_data %>%
  mutate(date = as.Date(date))

filtered_data <- filtered_data %>%
  mutate(date = as.Date(date))

# Specify the month and year for filtering
specific_month <- 7  # July
specific_year <- 2023  # Corrected from 2023 to 2022

# Filter the data for the specified month and year
gage_data_filtered <- gage_data %>%
  filter(month(date) == specific_month & year(date) == specific_year)

filtered_data_filtered <- merged_data %>%
  filter(month(date) == specific_month & year(date) == specific_year)
ggplot() +
  # Gage data
  geom_line(
    data = gage_data_filtered,
    aes(x = date, y = temperature, color = "Gage"),
    linewidth = 1,
    linetype = "dotted"
  ) +
  # TempEst data
  geom_line(
    data = filtered_data_filtered,
    aes(x = date, y = temp.mod, color = "TempEst"),
    linewidth = 1,
    linetype = "dotted"
  ) +
  scale_color_manual(
    values = c("Gage" = "darkgreen", "TempEst" = "orange"),
    name = "Source"
  ) +
  # Apply minimal theme with increased font sizes
  theme_minimal() +
  theme(
    plot.title = element_text(size = 22, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 16)
  ) +
  # Add labels
  labs(
    title = paste("Gage vs TempEst Temperature for", month.name[specific_month], specific_year),
    x = "Date",
    y = "Temperature (°C)"
  )
```
```{r}
precip_data <- read_csv("precipitation_data.csv")
precip_data <- precip_data %>%
  mutate(date = as.Date(date))
cloud_data <- cloud_data %>%
  mutate(date = as.Date(date))
specific_month <- 7    
specific_year <- 2023
precip_data_july <- precip_data %>%
  filter(month(date) == specific_month & year(date) == specific_year)
cloud_data_july <- cloud_data %>%
  filter(month(date) == specific_month & year(date) == specific_year)
combined_data <- left_join(precip_data_july, cloud_data_july, by = "date")
combined_long <- combined_data %>%
  select(date, precipitation, mean_cloud_cover) %>%
  pivot_longer(cols = c(precipitation, mean_cloud_cover), names_to = "Variable", values_to = "Value")

ggplot(combined_long, aes(x = date, y = Value)) +
  geom_line(linetype = "dotted") +
  facet_wrap(~ Variable, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 22, face = "bold"),
    strip.text = element_text(size = 18),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  ) +
  labs(
    title = paste("Precipitation and Mean Cloud Cover in", month.name[specific_month], specific_year),
    x = "Date",
    y = ""
  )
#correlation <- cor.test(combined_data$precipitation, combined_data$mean_cloud_cover, use = "complete.obs")
#print(correlation)
```

```{r}
# Filter the data for July 2023 and create date_no_year column
merged_data_weatherSR <- merged_data_weatherSR %>%
  filter(format(date, "%Y-%m") == "2022-07") %>%
  mutate(date_no_year = format(date, "%m-%d"))

# Step 1: Calculate the seasonal trend (mean temperature for each day of the year)
seasonal_trend <- merged_data_weatherSR %>%
  group_by(date_no_year) %>%
  summarise(mean_temp = mean(temp.mod, na.rm = TRUE))

# Step 2: Join seasonal trend back to the data and calculate temperature anomaly
merged_data_weatherSR <- merged_data_weatherSR %>%
  left_join(seasonal_trend, by = "date_no_year") %>%
  mutate(temp_anomaly = temp.mod - mean_temp)  # Calculate anomaly

# Step 3: Scale and normalize the temperature data
merged_data_weatherSR <- merged_data_weatherSR %>%
  group_by(date) %>%
  mutate(temp_scaled = (temp.mod - mean(temp.mod, na.rm = TRUE)) / sd(temp.mod, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(temp_normalized = (temp.mod - min(temp.mod, na.rm = TRUE)) /
                           (max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE)))

# View the first few rows to check
head(merged_data_weatherSR)
ggplot(merged_data_weatherSR, aes(x = weather_condition, y = temp_anomaly, fill = weather_condition)) +
  geom_boxplot() +
  labs(
    title = "Temperature Anomalies by Weather Condition in July 2023",
    x = "Weather Condition",
    y = "Temperature Anomaly (°C)"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(name = "Weather Condition")


```
# Create the categorical plot comparing temperature by weather condition
```{r}
merged_data_weatherSR <- merged_data_weatherSR %>%
  filter(
    date >= as.Date("2023-07-01") & date <= as.Date("2023-07-31"),
    !is.na(weather_condition)
  )
#box plot 
ggplot(merged_data_weatherSR, aes(x = weather_condition, y = temp.mod, fill = weather_condition)) +
  geom_boxplot() +
  scale_fill_viridis_d(name = "Weather Condition") +
  theme_minimal() +
  labs(
    title = "Temperature Comparison by Weather Condition in July 2023",
    x = "Weather Condition",
    y = "Temperature (°C)"
  )


```
```{r}

# Step 1: Filter data for July 2023
merged_data_weatherSR <- merged_data_weatherSR %>%
  filter(date >= as.Date("2022-07-01") & date <= as.Date("2022-12-31"))

temp_trend <- lm(temp.mod ~ date, data = merged_data_weatherSR)

merged_data_weatherSR <- merged_data_weatherSR %>%
  mutate(
    temp_detrended = temp.mod - predict(temp_trend)  # Subtract the predicted trend
  )
ggplot(merged_data_weatherSR, aes(x = date, y = temp_detrended, color = weather_condition)) +
  geom_point(size = 3, alpha = 0.7) +  
  scale_color_viridis_d(name = "Weather Condition") +
  theme_minimal() +
  labs(
    title = "Detrended Temperature Time Series Colored by Weather Condition (July 2023)",
    x = "Date",
    y = "Detrended Temperature (°C)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  
  )


```
# Excluding the TempEST 
```{r}
gage_dataSR <- gage_dataSR %>%
  mutate(date = as.Date(date))
gage_dataSR <- gage_dataSR %>%
  left_join(select(merged_data_weatherSR, date, weather_condition), by = "date")

specific_month <- 7  
specific_year <- 2022 

# Filter only the gage data for July 2023
gage_data_filtered <- gage_dataSR %>%
  filter(month(date) == specific_month & year(date) == specific_year)

gage_temp_trend <- lm(temperature ~ date, data = gage_data_filtered)

gage_data_filtered <- gage_data_filtered %>%
  mutate(temp_detrended = temperature - predict(gage_temp_trend))

ggplot(gage_data_filtered, aes(x = date, y = temp_detrended, color = weather_condition)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_d(name = "Weather Condition") +
  theme_minimal() +
  labs(
    title = paste("Detrended Temperature Time Series Colored by Weather Condition (July 2023)"),
    x = "Date",
    y = "Detrended Temperature (°C)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  
  )

```
#Annual Plot For TempEST, Cloud Cover, Precipitation
```{r}
merged_data_weatherSR <- merged_data_weatherSR %>%
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2022-12-31"))

ggplot(merged_data_weatherSR, aes(x = date, y = temp.mod, color = weather_condition)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_color_viridis_d(name = "Weather Condition") +
  theme_minimal() +
  labs(
    title = "Temperature Time Series Colored by Weather Condition (2022)",
    x = "Date",
    y = "Modeled Temperature (°C)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
#Annual excluding TempEST
```{r}
gage_dataSR <- gage_dataSR %>%
  mutate(date = as.Date(date)) %>%
  left_join(select(merged_data_weatherSR, date, weather_condition), by = "date")

gage_data_filtered <- gage_dataSR %>%
  filter(year(date) == 2022)

# Plot raw temperature, no detrending
ggplot(gage_data_filtered, aes(x = date, y = temperature, color = weather_condition)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_d(name = "Weather Condition") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  theme_minimal() +
  labs(
    title = "Temperature Time Series Colored by Weather Condition (2022, Gage Data)",
    x = "Date",
    y = "Temperature (°C)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
# Adds the precipitation and cloud cover conditioning column(run before interpolation and scatter plot)
```{r}
# Load and preprocess precipitation data
precipitation_dataSR <- read_csv("precipitation_data.csv") %>%
  mutate(date = as.Date(date)) %>%
  group_by(date) %>%
  summarise(precipitation = sum(precipitation, na.rm = TRUE))
# Join precipitation data with merged data
merged_data_weatherSR <- merged_dataSR %>%
  left_join(precipitation_dataSR, by = "date")
# Define thresholds
precip_threshold <- 0
cloud_threshold <- 30

# Add weather condition column
merged_data_weatherSR <- merged_data_weatherSR %>%  # Use merged_data_weather here
  mutate(
    weather_condition = case_when(
      precipitation > precip_threshold & mean_cloud_cover > cloud_threshold ~ "Raining and Cloudy",
      precipitation <= precip_threshold & mean_cloud_cover > cloud_threshold ~ "Cloudy",
      mean_cloud_cover <= cloud_threshold ~ "No Clouds or Rain",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(weather_condition = factor(weather_condition, 
                                    levels = c("No Clouds or Rain", "Cloudy", "Raining and Cloudy")))
                                  
```

```{r}
# Load and preprocess precipitation data
precipitation_data4 <- read_csv("precipitation_dataSR.csv") %>%
  mutate(date = as.Date(date)) %>%
  group_by(date) %>%
  summarise(precipitation = sum(precipitation, na.rm = TRUE))
# Join precipitation data with merged data
merged_data_weather4 <- merged_data4 %>%
  left_join(precipitation_data2, by = "date")
# Define thresholds
precip_threshold <- 0
cloud_threshold <- 30

# Add weather condition column
merged_data_weather4 <- merged_data_weather4 %>%  # Use merged_data_weather here
  mutate(
    weather_condition = case_when(
      precipitation > precip_threshold & mean_cloud_cover > cloud_threshold ~ "Raining and Cloudy",
      precipitation <= precip_threshold & mean_cloud_cover > cloud_threshold ~ "Cloudy",
      mean_cloud_cover <= cloud_threshold ~ "No Clouds or Rain",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(weather_condition = factor(weather_condition, 
                                    levels = c("No Clouds or Rain", "Cloudy", "Raining and Cloudy")))
          
```

#Scatter quantile plot with gage data
``` {r }
gage_dataSR <- read_csv('PotomacDataGage.csv')
gage_dataSR <- gage_dataSR %>%
  mutate(date = as.Date(date)) %>%
  left_join(select(merged_data_weatherSR, date, weather_condition), by = "date")
gage_data_filtered <- gage_dataSR %>%
  filter(year(date) == 2022) %>%
  mutate(
    day_of_year = yday(date),
    weather_condition = case_when(
      is.na(weather_condition)                ~ "No Cloud/Rain",
      weather_condition == "No Clouds or Rain" ~ "No Cloud/Rain",
      weather_condition == "Cloudy"            ~ "Cloudy",
      weather_condition == "Raining and Cloudy"~ "Cloudy",   
      TRUE                                    ~ as.character(weather_condition)
    )
  )
river_name <- "Potomac River"  
plot_year <- unique(lubridate::year(gage_data_filtered$date))
plot_year <- unique(lubridate::year(gage_data_filtered$date))
# Reshape gage_summary for wide format to compare q50 for both conditions by day
gage_summary_wide <- gage_summary %>%
  filter(weather_condition %in% c("Cloudy", "No Cloud/Rain")) %>%
  select(day_of_year, weather_condition, q50) %>%
  tidyr::pivot_wider(names_from = weather_condition, values_from = q50)
#medians
cloudy_median <- median(gage_summary$q50[gage_summary$weather_condition == "Cloudy"], na.rm = TRUE)
nocloud_median <- median(gage_summary$q50[gage_summary$weather_condition == "No Cloud/Rain"], na.rm = TRUE)
diff_median <- cloudy_median - nocloud_median

# Calculate R-squared for median temps on same days
matching_days <- na.omit(gage_summary_wide)
perf_metric <- cor(matching_days$Cloudy, matching_days$`No Cloud/Rain`)^2
#cohen d analysis

cohen_d <- cohen.d(
  gage_data_filtered$temperature[gage_data_filtered$weather_condition == "Cloudy"],
  gage_data_filtered$temperature[gage_data_filtered$weather_condition == "No Cloud/Rain"]
)
ci_lower <- cohen_d$conf.int[1]
ci_upper <- cohen_d$conf.int[2]

gage_summary <- gage_data_filtered %>%
  group_by(day_of_year, weather_condition) %>%
  summarise(
    q10 = quantile(temperature, 0.10, na.rm = TRUE),
    q25 = quantile(temperature, 0.25, na.rm = TRUE),
    q50 = quantile(temperature, 0.50, na.rm = TRUE),
    q75 = quantile(temperature, 0.75, na.rm = TRUE),
    q90 = quantile(temperature, 0.90, na.rm = TRUE),
    .groups = "drop"
  )
gage_summary_long <- gage_summary %>%
  pivot_longer(
    cols = starts_with("q"),
    names_to = "quantile",
    values_to = "temperature"
  )

ggplot(gage_summary, aes(x = day_of_year)) +
  # 80% quantile band
  geom_ribbon(aes(ymin = q10, ymax = q90, fill = weather_condition), alpha = 0.3) +
  # 50% quantile band
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = weather_condition), alpha = 0.5) +
  # Median line
  geom_line(aes(y = q50, color = weather_condition), size = 1.2) +
  # Quantile scatter points
  geom_point(
    data = gage_summary_long,
    aes(y = temperature, shape = quantile, color = weather_condition),
    size = 1.2,
    position = position_jitter(width = 0.8, height = 0) 
  ) +
    scale_fill_manual(values = c(
    "Cloudy" = "gold",
    "Clear" = "gray70",
    "No Cloud/Rain" = "gray60"
  )) +
  scale_color_manual(values = c(
    "Cloudy" = "orange3",
    "Clear" = "black",
    "No Cloud/Rain" = "gray30"
  )) +

  scale_shape_manual(values = c(
    "q10" = 1, 
    "q25" = 2, 
    "q50" = 16, 
    "q75" = 17, 
    "q90" = 15)) +
  scale_x_continuous(
    breaks = seq(0, 366, by = 90),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(
      floor(min(gage_summary_long$temperature, na.rm = TRUE)),
      ceiling(max(gage_summary_long$temperature, na.rm = TRUE)),
      by = 5
    ),
    expand = c(0, 0)
  ) +
  
  labs(
  title = sprintf("Daily Temperature Quantiles by Weather Condition (%s, %s)", river_name, plot_year),
  subtitle = sprintf(
  "Median Temp R² (Cloudy vs. No Cloud/Rain): %s\nMedian Temp Difference (Cloudy - No Cloud/Rain): %.2f°C\nCohen's d: %.2f (95%% CI: %.2f, %.2f) ",
  ifelse(is.na(perf_metric), "NA", sprintf("%.2f", perf_metric)),
  diff_median,
  cohen_d$estimate, ci_lower, ci_upper
),

  x = "Day of Year",
  y = "Temperature (°C)",
  color = "Weather Condition",
  fill = "Weather Condition",
  shape = "Quantile"
)+
  theme_minimal()
```


# Prediction with a Trained Model

TempEst 2 comes with a pretrained model, `model.rda`.  Loading this RData file
will create a `model` function.  This function takes in the predictor data frame
and adds the columns `temp.doy` (day-of-year mean temperature), `temp.anom`
(daily anomaly relative to day-of-year mean, i.e., de-seasoned), and `temp.mod`
(overall modeled temperature for each day, the sum of the previous two).

```{r message=FALSE}
load("model.rda")
predSR <- model(dataSR) %>% drop_na()
```
```{r}
load("model.rda")
pred2 <- model(data2) %>% drop_na()
load("model.rda")
pred3 <- model(data3) %>% drop_na()
load("model.rda")
pred4 <- model(data4) %>% drop_na()
```
```{r}
predSR <- predSR %>%
  mutate(id = as.numeric(id), date = as.Date(date))
cloud_dataSR <- cloud_dataSR%>%
  mutate(id = as.numeric(gsub("Point", "", point_id)), 
    date = as.Date(date, format = "%Y-%m-%d")) %>%
  select(-point_id)
merged_dataSR <- left_join(predSR, cloud_dataSR, by = c("id", "date"))
head(merged_dataSR)
```

```{r}
pred2 <- pred2 %>%
  mutate(id = as.numeric(id), date = as.Date(date))
cloud_data2 <- cloud_data2%>%
  mutate(id = as.numeric(gsub("Point", "", point_id)), 
    date = as.Date(date, format = "%Y-%m-%d")) %>%
  select(-point_id)
merged_data2 <- left_join(pred2, cloud_data2, by = c("id", "date"))
head(merged_data2)

pred3 <- pred3 %>%
  mutate(id = as.numeric(id), date = as.Date(date))
cloud_data3 <- cloud_data3%>%
  mutate(id = as.numeric(gsub("Point", "", point_id)), 
    date = as.Date(date, format = "%Y-%m-%d")) %>%
  select(-point_id)
merged_data3 <- left_join(pred3, cloud_data3, by = c("id", "date"))
head(merged_data3)

pred4 <- pred4 %>%
  mutate(id = as.numeric(id), date = as.Date(date))
cloud_data4 <- cloud_data4%>%
  mutate(id = as.numeric(gsub("Point", "", point_id)), 
    date = as.Date(date, format = "%Y-%m-%d")) %>%
  select(-point_id)
merged_data4 <- left_join(pred4, cloud_data4, by = c("id", "date"))
head(merged_data4)
```
#interpolation for single river(based off TempEst2)
```{r}

df_keyed <- merged_data_weatherSR %>%
  mutate(station_key = paste0(round(lat.x, 4), "_", round(lon.x, 4)))

station_coords <- df_keyed %>%
  filter(!is.na(lat.x) & !is.na(lon.x)) %>%
  group_by(station_key) %>%
  summarise(
    lat.x = mean(lat.x, na.rm = TRUE),
    lon.x = mean(lon.x, na.rm = TRUE),
    .groups = "drop"
  )

df_daily <- df_keyed %>%
  group_by(station_key, date) %>%
  summarise(
    lst      = mean(lst, na.rm = TRUE),
    humidity = mean(humidity, na.rm = TRUE),
    temp.mod = mean(temp.mod, na.rm = TRUE),
    .groups = "drop"
  )

df_complete <- df_daily %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  arrange(station_key, date)
df_interp <- df_complete %>%
  group_by(station_key) %>%
  arrange(date) %>%
  # store original in a new column
  mutate(temp_orig = temp.mod,
         lst_orig = lst,
         humidity_orig = humidity) %>%
  # do interpolation into separate cols
  mutate(
    lst_interp = na.approx(lst, x = as.numeric(date), na.rm = FALSE, rule = 2),
    humidity_interp = na.approx(humidity, x = as.numeric(date), na.rm = FALSE, rule = 2),
    temp_interp = na.approx(temp.mod, x = as.numeric(date), na.rm = FALSE, rule = 2)
  ) %>%
  # now choose original where available, else interpolated
  mutate(
    lst      = if_else(!is.na(lst_orig), lst_orig, lst_interp),
    humidity = if_else(!is.na(humidity_orig), humidity_orig, humidity_interp),
    temp.mod = if_else(!is.na(temp_orig), temp_orig, temp_interp)
  ) %>%
  ungroup() %>%
  select(-ends_with("_orig"), -ends_with("_interp"))

# then re-join coords
df_interp <- df_interp %>%
  left_join(station_coords, by = "station_key")



```
#interpolation for four rivers in one binded stack(based off TempEst2)
```{r}
merged_all <- bind_rows(
  merged_data_weatherSR,
  merged_data_weather2,
  merged_data_weather3,
  merged_data_weather4
)
df_keyed_all <- merged_all %>%
  mutate(station_key = paste0(round(lat.x, 4), "_", round(lon.x, 4)))

station_coords <- df_keyed %>%
  filter(!is.na(lat.x) & !is.na(lon.x)) %>%
  group_by(station_key) %>%
  summarise(
    lat.x = mean(lat.x, na.rm = TRUE),
    lon.x = mean(lon.x, na.rm = TRUE),
    .groups = "drop"
  )

df_daily <- df_keyed %>%
  group_by(station_key, date) %>%
  summarise(
    lst      = mean(lst, na.rm = TRUE),
    humidity = mean(humidity, na.rm = TRUE),
    temp.mod = mean(temp.mod, na.rm = TRUE),
    .groups = "drop"
  )

df_complete <- df_daily %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  arrange(station_key, date)
df_interp_all <- df_complete %>%
  group_by(station_key) %>%
  arrange(date) %>%
  # store original in a new column
  mutate(temp_orig = temp.mod,
         lst_orig = lst,
         humidity_orig = humidity) %>%
  # do interpolation into separate cols
  mutate(
    lst_interp = na.approx(lst, x = as.numeric(date), na.rm = FALSE, rule = 2),
    humidity_interp = na.approx(humidity, x = as.numeric(date), na.rm = FALSE, rule = 2),
    temp_interp = na.approx(temp.mod, x = as.numeric(date), na.rm = FALSE, rule = 2)
  ) %>%
  # now choose original where available, else interpolated
  mutate(
    lst      = if_else(!is.na(lst_orig), lst_orig, lst_interp),
    humidity = if_else(!is.na(humidity_orig), humidity_orig, humidity_interp),
    temp.mod = if_else(!is.na(temp_orig), temp_orig, temp_interp)
  ) %>%
  ungroup() %>%
  select(-ends_with("_orig"), -ends_with("_interp"))

# then re-join coords
df_interp_all <- df_interp_all %>%
  left_join(station_coords, by = "station_key")

```
#interpolation based off gage
```{r}

gage_df <- gage_dataSR %>%
  mutate(
    station_key = id,              # use the gage ID as station_key
    date        = as.Date(date)
  )

# Pull unique IDs and fetch their coordinates
unique_ids <- unique(gage_df$id)
gage_meta  <- readNWISsite(unique_ids) %>%
  select(id = site_no, lat.x = dec_lat_va, lon.x = dec_long_va)

# Join coords back to the gage series
gage_keyed <- gage_df %>%
  left_join(gage_meta, by = "id")

# Collapse to one daily value per station_key
df_daily_gage <- gage_keyed %>%
  group_by(station_key, date) %>%
  summarize(
    temp.mod = mean(temperature, na.rm = TRUE),
    .groups   = "drop"
  )

# Complete the full date sequence per gauge
df_complete_gage <- df_daily_gage %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  ungroup()

# Interpolate missing days, preserving originals
df_interp_gage <- df_complete_gage %>%
  group_by(station_key) %>%
  arrange(date) %>%
  mutate(
    temp_orig   = temp.mod,
    temp_interp = na.approx(
                    temp.mod,
                    x      = as.numeric(date),
                    na.rm  = FALSE,
                    rule   = 2
                  ),
    temp.mod    = if_else(!is.na(temp_orig), temp_orig, temp_interp)
  ) %>%
  ungroup() %>%
  select(-temp_orig, -temp_interp)

# Reattach the coordinates
df_interp_gage <- df_interp_gage %>%
  left_join(
    gage_meta %>% rename(station_key = id),
    by = "station_key"
  )

# Inspect the result
glimpse(df_interp_gage)

```
#gage interpolation(where it is only na) vs tempest2(only interpolated when it is needed too)
```{r}



# Build your grid and gauge location tables:

grid_pts <- df_interp %>%
  distinct(grid_key  = station_key,
           lat.grid  = lat.x,
           lon.grid  = lon.x)

# Suppose you pulled your gauge metadata like this:
gage_meta <- readNWISsite(unique(gage_dataSR$id)) %>%
  select(
    station_key = site_no,
    lat_gage    = dec_lat_va,
    lon_gage    = dec_long_va
  )

# 1. Map each gauge to its nearest grid point:
targets <- gage_meta %>%
  rowwise() %>%
  mutate(
    grid_key = grid_pts$grid_key[
      which.min(
        distHaversine(
          c(lon_gage, lat_gage),
          cbind(grid_pts$lon.grid, grid_pts$lat.grid)
        )
      )
    ]
  ) %>%
  ungroup() %>%
  select(station_key, grid_key)

# 2. Prepare the two time series:

# Gauge-interpolated series (from your hold‐out pipeline)
gage_ts <- df_interpolated %>%
  rename(temp_gage = temp_pred) %>%
  select(station_key, date, temp_gage)

# TempEST2 grid series
model_ts <- df_interp %>%
  rename(
    grid_key   = station_key,
    temp_model = temp.mod
  ) %>%
  select(grid_key, date, temp_model)

# 3. Join them all together by matching gauge→grid and by date:
comparison <- targets %>%
  inner_join(gage_ts,  by = "station_key") %>%    # gauge series
  inner_join(model_ts, by = c("grid_key", "date"))  # model series

# 4. Compute summary statistics:

# Per‐gage
gage_summary <- comparison %>%
  group_by(station_key) %>%
  summarize(
    mean_bias = mean(temp_model - temp_gage, na.rm = TRUE),
    RMSE      = rmse(temp_gage, temp_model),
    MAE       = mae(temp_gage, temp_model),
    n_days    = n(),
    .groups   = "drop"
  )

print(gage_summary)

# Overall
overall_summary <- gage_summary %>%
  summarize(
    avg_bias = mean(mean_bias, na.rm = TRUE),
    avg_RMSE = mean(RMSE,        na.rm = TRUE),
    avg_MAE  = mean(MAE,         na.rm = TRUE)
  )

print(overall_summary)

# 5. Plot for one example gauge:
demo_id <- '09085000'

comparison %>%
  filter(station_key == demo_id) %>%
  gather(source, temp, temp_gage, temp_model) %>%
  ggplot(aes(date, temp, color = source)) +
    geom_line() +
    theme_bw() +
    labs(
      title = paste("Gauge vs Model for", demo_id),
      x     = "Date",
      y     = "Temperature (°C)"
    )



```

```{r}
library(dplyr)
library(tidyr)
library(zoo)

# 1. Build your daily gauge series and complete dates
df_complete_gage <- gage_dataSR %>%
  mutate(
    station_key = id,
    date        = as.Date(date)
  ) %>%
  group_by(station_key, date) %>%
  summarize(temp.mod = mean(temperature, na.rm = TRUE), .groups = "drop") %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  ungroup()

# 2. “Blank” temp.mod but keep a copy in temp_orig
df_blank <- df_complete_gage %>%
  mutate(
    temp_orig = temp.mod,
    temp.mod  = NA_real_
  )

# 3. Interpolate using the ORIGINAL values (temp_orig) as anchors
df_interp_all_na <- df_blank %>%
  group_by(station_key) %>%
  arrange(date) %>%
  mutate(
    temp.mod = na.approx(
      temp_orig,               # interpolate the original vector
      x         = as.numeric(date),
      na.rm     = FALSE,
      rule      = 2
    )
  ) %>%
  ungroup() %>%
  select(-temp_orig)

# 4. Compare
df_compare <- df_interp_all_na %>%
  rename(temp_filled = temp.mod) %>%
  left_join(
    df_complete_gage %>% 
      rename(temp_true = temp.mod),
    by = c("station_key","date")
  )

glimpse(df_compare)

```
#plot for gage interpolation where values became na then interpolation happens
```{r}


# Collapse to one raw daily value per gauge
df_daily <- gage_dataSR %>%
  mutate(
    station_key = id,
    date        = as.Date(date)
  ) %>%
  group_by(station_key, date) %>%
  summarize(
    temp_true = mean(temperature, na.rm = TRUE),
    .groups   = "drop"
  )

# Complete the calendar (introduce explicit NAs)
df_complete <- df_daily %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  ungroup()

# Linear‐interpolate only the gaps, preserving originals
df_interp_gage <- df_complete %>%
  group_by(station_key) %>%
  arrange(date) %>%
  mutate(
    temp_orig   = temp_true,                              # save the raw
    temp.mod    = na.approx(                              # interp the raw vector
                    temp_true,
                    x      = as.numeric(date),
                    na.rm  = FALSE,
                    rule   = 2
                 ),
    temp.mod    = if_else(!is.na(temp_orig),             # keep raw where it exists
                          temp_orig,
                          temp.mod)
  ) %>%
  ungroup() %>%
  select(station_key, date, temp_true, temp_interp = temp.mod)

# Choose one gauge to examine
demo_gage <- "09085000"
raw  <- df_daily  %>% filter(station_key == demo_gage)
interp <- df_interp_gage %>% filter(station_key == demo_gage)

# Plot: blue line = full interpolated, red points = raw observations
ggplot() +
  geom_line(data = interp,
            aes(x = date, y = temp_interp),
            color = "steelblue", size = 0.7) +
  geom_point(data = raw,
             aes(x = date, y = temp_true),
             color = "firebrick", size = 1.2, alpha = 0.8) +
  theme_bw() +
  labs(
    x     = "Date",
    y     = "Temperature (°C)",
    title = paste("Gauge", demo_gage, "— Raw vs. Interpolated"),
    subtitle = "Blue = gap‐filled daily series; Red dots = true observations"
  )
library(Metrics)

# Use df_compare (raw vs interpolated on days you actually observed)
#    It should have: station_key, date, temp_true, temp_interp
glimpse(df_compare)

# Compute per-gauge error metrics
error_metrics <- df_compare %>%
  group_by(station_key) %>%
  summarize(
    RMSE = rmse(temp_true, temp_interp),
    MAE  = mae(temp_true, temp_interp),
    .groups = "drop"
  )

print(error_metrics)

# Overall stats
overall <- error_metrics %>%
  summarize(
    avg_RMSE = mean(RMSE, na.rm = TRUE),
    avg_MAE  = mean(MAE,  na.rm = TRUE)
  )

print(overall)

# Optionally append to your summary table for the ‘Interpolated’ series row:
summary_stats <- summary_stats %>%
  mutate(
    RMSE = if_else(Series == "Interpolated", overall$avg_RMSE, NA_real_),
    MAE  = if_else(Series == "Interpolated", overall$avg_MAE,  NA_real_)
  )

print(summary_stats)







```
#gage interpolation making every value na then interpolating based off original values, this one works
```{r}


df_complete <- gage_dataSR %>%
  mutate(
    station_key = id,
    date        = as.Date(date)
  ) %>%
  group_by(station_key, date) %>%
  summarize(
    temp.mod = mean(temperature, na.rm = TRUE),  # <-- temp.mod comes back
    .groups   = "drop"
  ) %>%
  group_by(station_key) %>%
  complete(
    date = seq(min(date), max(date), by = "day")
  ) %>%
  ungroup()
# Select a 10% hold-out set to act as our test data
set.seed(42) # for reproducible results
validation_set <- df_complete %>%
  filter(!is.na(temp.mod)) %>%         # Only pick from days we have data for
  group_by(station_key) %>%
  slice_sample(prop = 0.10) %>%      # Randomly sample 10% of rows from each gage
  ungroup() %>%
  select(station_key, date, temp_true = temp.mod) # Keep the true values

# Mask the hold-out values in the main dataframe
df_masked <- df_complete %>%
  # If a row is in our validation_set, make its temp NA; otherwise, keep it
  mutate(temp_for_interp = if_else(
    date %in% validation_set$date[validation_set$station_key == station_key],
    NA_real_,
    temp.mod
  ))

# Perform the interpolation on the masked data
df_interpolated <- df_masked %>%
  group_by(station_key) %>%
  arrange(date) %>%
  mutate(
    temp_pred = na.approx(temp_for_interp, na.rm = FALSE, rule = 2)
  ) %>%
  ungroup()

# Join the interpolated values back to the true values for comparison
final_comparison <- validation_set %>%
  # Grab the newly predicted values for our test locations
  left_join(
    df_interpolated %>% select(station_key, date, temp_pred),
    by = c("station_key", "date")
  ) %>%
  # Clean up any rows where interpolation wasn't possible 
  filter(!is.na(temp_pred))

print(
  paste0(
    "Overall RMSE: ",
    round(rmse(final_comparison$temp_true, final_comparison$temp_pred), 2),
    " °C"
  )
)

ggplot(final_comparison, aes(x = temp_true, y = temp_pred)) +
  geom_abline(linetype = "dashed", color = "red", linewidth = 1) +
  geom_point(alpha = 0.7) +
  theme_bw() +
  labs(
    title = "Interpolated vs. Observed Gage Temperatures",
    subtitle = "Comparing interpolated values to true, raw observations",
    x = "Observed, Raw Temperature (°C)",
    y = "Interpolated Temperature (°C)"
  ) +
  coord_fixed() 
# Create a summary statistics table
final_comparison <- final_comparison %>%
  mutate(error = temp_pred - temp_true)

# Create a summary table for each individual gage
summary_stats_by_gage <- final_comparison %>%
  group_by(station_key) %>%
  summarize(
    n_obs       = n(),
    mean_bias   = mean(error),
    mae         = mae(temp_true, temp_pred),
    rmse        = rmse(temp_true, temp_pred),
    r_squared   = cor(temp_true, temp_pred)^2,
    .groups     = "drop"
  )

# Create an overall summary table by averaging the per-gage stats
overall_summary <- summary_stats_by_gage %>%
  summarize(
    avg_mae       = mean(mae),
    avg_rmse      = mean(rmse),
    avg_r_squared = mean(r_squared)
  )

# Print the summary tables
print("--- Summary Statistics by Gage ---")
print(summary_stats_by_gage)

print("--- Overall Model Performance ---")
print(overall_summary)

```
```{r}
df_complete <- gage_dataSR %>%
  mutate(
    station_key = id,
    date        = as.Date(date)
  ) %>%
  group_by(station_key, date) %>%
  summarize(temp.mod = mean(temperature, na.rm = TRUE), .groups = "drop") %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  ungroup()

set.seed(42)
validation_set <- df_complete %>%
  filter(!is.na(temp.mod)) %>%
  group_by(station_key) %>%
  slice_sample(prop = 0.10) %>%
  ungroup() %>%
  select(station_key, date, temp_true = temp.mod)

df_interpolated <- df_complete %>%
  mutate(
    temp_for_interp = if_else(
      station_key %in% validation_set$station_key &
      date %in% validation_set$date[validation_set$station_key == station_key],
      NA_real_, temp.mod
    )
  ) %>%
  group_by(station_key) %>%
  arrange(date) %>%
  mutate(temp_pred = na.approx(temp_for_interp, x = as.numeric(date), na.rm = FALSE, rule = 2)) %>%
  ungroup()

final_comparison <- validation_set %>%
  left_join(df_interpolated %>% select(station_key, date, temp_pred),
            by = c("station_key","date")) %>%
  filter(!is.na(temp_pred)) %>%
  mutate(error = temp_pred - temp_true)

# Prepare gage and grid locations

# Grid locations from your TempEST2 output
grid_pts <- df_interp %>%
  distinct(grid_key = station_key, lat.x, lon.x)

# Gage locations via USGS metadata
gage_ids  <- unique(gage_dataSR$id)
gage_meta <- readNWISsite(gage_ids) %>%
  select(station_key = site_no,
         lat.gage     = dec_lat_va,
         lon.gage     = dec_long_va)

#  Map each gage to its nearest grid point

gauge_to_grid <- gage_meta %>%
  rowwise() %>%
  mutate(
    grid_key = grid_pts$grid_key[
      which.min(distHaversine(
        c(lon.gage, lat.gage),
        cbind(grid_pts$lon.x, grid_pts$lat.x)
      ))
    ]
  ) %>%
  ungroup() %>%
  select(station_key, grid_key)

# Combine gage‐interpolated & TempEST2 for direct comparison

comparison <- df_interpolated %>%
  rename(temp_gage = temp_pred) %>%
  left_join(gauge_to_grid, by = "station_key") %>%
  inner_join(
    df_interp %>%
      rename(grid_key   = station_key,
             temp_model = temp.mod),
    by = c("grid_key","date")
  ) %>%
  filter(!is.na(temp_gage), !is.na(temp_model)) %>%
  mutate(error_gm = temp_model - temp_gage)

#  Compute and print error metrics

overall_rmse <- rmse(comparison$temp_gage, comparison$temp_model)
overall_mae  <- mae (comparison$temp_gage, comparison$temp_model)

per_gage_metrics <- comparison %>%
  group_by(station_key) %>%
  summarize(
    n       = n(),
    RMSE    = rmse(temp_gage, temp_model),
    MAE     = mae (temp_gage, temp_model),
    bias    = mean(temp_model - temp_gage),
    .groups = "drop"
  )

cat("Overall Gage vs TempEST2 RMSE:", round(overall_rmse,2), "°C\n")
cat("Overall Gage vs TempEST2 MAE :", round(overall_mae,2),  "°C\n")
print(per_gage_metrics)

# Plot gage vs. TempEST2 for each station

ggplot(comparison, aes(x = temp_gage, y = temp_model)) +
  geom_abline(lty = "dashed", color = "gray50") +
  geom_point(alpha = 0.5) +
  facet_wrap(~ station_key, scales = "free", ncol = 2) +
  theme_bw() +
  labs(
    x        = "Gage‐Interpolated Temp (°C)",
    y        = "TempEST2 Model Temp (°C)",
    title    = "Gage vs. TempEST2 Interpolation Comparison",
    subtitle = sprintf("Overall RMSE = %.2f °C, MAE = %.2f °C",
                       overall_rmse, overall_mae)
  )


```
```{r}
library(ggplot2)

ggplot(comparison, aes(x = temp_gage, y = temp_model)) +
  # add a clean 1:1 line
  geom_abline(intercept = 0, slope = 1, color = "#333333", size = 1, linetype = "solid") +
  # show point density with semi‐transparent points
  geom_point(alpha = 0.3, shape = 16) +
  # overlay a LOESS smoother to show systematic bias
  stat_smooth(method = "loess", color = "#D55E00", size = 1, se = FALSE) +
  # facet per station
  facet_wrap(~ station_key, scales = "free", ncol = 2) +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "#F0F0F0", color = NA),
    panel.grid.minor  = element_blank()
  ) +
  labs(
    x        = "Gage-Interpolated Temp (°C)",
    y        = "TempEST2 Model Temp (°C)",
    title    = "Gage vs. TempEST2: Predicted vs. Observed",
    subtitle = sprintf(
      "Overall RMSE = %.2f °C, MAE = %.2f °C",
      overall_rmse, overall_mae
    ),
    caption  = "Gray line = perfect agreement; orange curve = LOESS fit"
  )

```

#single day validation for TempEst2 interpolation
```{r}

# Choose the date you want to test
selected_date <- as.Date("2022-07-31 ")

df_keyed <- merged_data_weatherSR %>%
  filter(!is.na(lat.x), !is.na(lon.x), !is.na(date)) %>%
  mutate(station_key = paste0(round(lat.x,4), "_", round(lon.x,4)))

station_coords <- df_keyed %>%
  group_by(station_key) %>%
  summarize(lat.x = mean(lat.x), lon.x = mean(lon.x), .groups="drop")

df_daily <- df_keyed %>%
  group_by(station_key, date) %>%
  summarize(temp.mod = mean(temp.mod, na.rm=TRUE), .groups="drop")

df_complete <- df_daily %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by="day")) %>%
  ungroup()

df_masked <- df_complete %>%
  mutate(
    temp_for_interp = if_else(date == selected_date, NA_real_, temp.mod)
  )

df_interp_masked <- df_masked %>%
  group_by(station_key) %>%
  arrange(date) %>%
  mutate(
    temp_pred = na.approx(
      temp_for_interp,
      x    = as.numeric(date),
      na.rm = FALSE,
      rule  = 2
    )
  ) %>%
  ungroup() %>%
  # bring back coords if desired
  left_join(station_coords, by="station_key")

df_true <- merged_data_weatherSR %>%
  filter(date == selected_date, !is.na(temp.mod)) %>%
  mutate(station_key = paste0(round(lat.x,4), "_", round(lon.x,4))) %>%
  select(station_key, date, temp_true = temp.mod)

df_compare <- df_interp_masked %>%
  filter(date == selected_date) %>%
  inner_join(df_true, by=c("station_key","date")) %>%
  filter(!is.na(temp_true), !is.na(temp_pred))

if(nrow(df_compare)==0) stop("No stations to compare on ", selected_date)
rmse_val <- rmse(df_compare$temp_true, df_compare$temp_pred)
mae_val  <- mean(abs(df_compare$temp_true - df_compare$temp_pred))

cat("Interpolation RMSE on", selected_date, ":", round(rmse_val,3), "°C\n",
    "Interpolation MAE  on", selected_date, ":", round(mae_val,3), "°C\n")

```
# validating the interpolation
```{r}
set.seed(42)

# prepare the data
df_keyed <- merged_data_weatherSR %>%
  filter(!is.na(lat.x), !is.na(lon.x), !is.na(date)) %>%
  mutate(station_key = paste0(round(lat.x,4), "_", round(lon.x,4)))

station_coords <- df_keyed %>%
  group_by(station_key) %>%
  summarize(lat.x = mean(lat.x), lon.x = mean(lon.x), .groups="drop")

df_daily <- df_keyed %>%
  group_by(station_key, date) %>%
  summarize(temp.mod = mean(temp.mod, na.rm=TRUE), .groups="drop")

df_complete <- df_daily %>%
  group_by(station_key) %>%
  complete(date = seq(min(date), max(date), by="day")) %>%
  ungroup()

# 2. Identify all (station_key, month) combos
month_list <- df_daily %>%
  filter(!is.na(temp.mod)) %>%
  mutate(month = floor_date(date, "month")) %>%
  distinct(station_key, month)

# 3. Sample up to 3 months per station
cv_months <- month_list %>%
  group_by(station_key) %>%
  reframe(
    month = sample(month, size = min(3, n()), replace = FALSE)
  ) %>%
  ungroup()


# 4. returns metrics and identifiers
evaluate_month <- function(st_key, mnth) {
  df_masked <- df_complete %>%
    filter(station_key == st_key) %>%
    mutate(
      temp_for_interp = if_else(floor_date(date, "month") == mnth,
                                NA_real_, temp.mod)
    )
  df_pred <- df_masked %>%
    arrange(date) %>%
    mutate(
      temp_pred = na.approx(temp_for_interp,
                            x    = as.numeric(date),
                            na.rm = FALSE,
                            rule  = 2)
    )
  df_true <- df_daily %>%
    filter(station_key == st_key,
           floor_date(date, "month") == mnth) %>%
    select(date, temp_true = temp.mod)
  df_comp <- inner_join(df_pred, df_true, by = "date")
  tibble(
    station_key = st_key,
    month        = mnth,
    rmse         = rmse(df_comp$temp_true, df_comp$temp_pred),
    mae          = mean(abs(df_comp$temp_true - df_comp$temp_pred), na.rm = TRUE),
    n_days       = nrow(df_comp)
  )
}

# 5. Loop with purrr::pmap_dfr
cv_results <- purrr::pmap_dfr(
  list(cv_months$station_key, cv_months$month),
  evaluate_month
)

# 6. summarize
cv_results <- cv_results %>%
  left_join(station_coords, by = "station_key")

summary_stats <- cv_results %>%
  summarize(
    mean_RMSE = mean(rmse, na.rm=TRUE),
    sd_RMSE   = sd(rmse,   na.rm=TRUE),
    mean_MAE  = mean(mae,  na.rm=TRUE),
    sd_MAE    = sd(mae,    na.rm=TRUE)
  )

print(cv_results)     # per‐station/month metrics
print(summary_stats)  # overall performance



```
#validating the interpolation with usgs gages
```{r}
gage_ids <- gage_dataSR %>% distinct(id) %>% pull(id)

#get site metadata (also lat/lon)
site_meta <- readNWISsite(gage_ids) %>%
  select(id = site_no, lat_gage = dec_lat_va, lon_gage = dec_long_va)

# merge coordinates into your gage table
gage_with_coords <- gage_dataSR %>%
  mutate(date = as.Date(date)) %>%
  left_join(site_meta, by = "id")

# select the date
selected_date <- as.Date("2022-07-10")

interp_day <- df_full_interp %>%
  filter(date == selected_date) %>%
  select(station_key, temp_pred, lat.x, lon.x)

# for each gage reading on that date, find the nearest grid point
gage_matched <- gage_with_coords %>%
  filter(date == selected_date, !is.na(temperature)) %>%
  rowwise() %>%
  mutate(
    nearest_station = interp_day$station_key[
      which.min(distHaversine(
        c(lon_gage, lat_gage),
        cbind(interp_day$lon.x, interp_day$lat.x)
      ))
    ],
    temp_pred = interp_day$temp_pred[which(interp_day$station_key == nearest_station)]
  ) %>%
  ungroup()

# compare observed vs. interpolated
rmse_val <- rmse(gage_matched$temperature, gage_matched$temp_pred)
mae_val  <- mae(gage_matched$temperature, gage_matched$temp_pred)

cat("Gauge vs. Interp RMSE:", round(rmse_val, 3), "°C\n",
    "Gauge vs. Interp MAE :", round(mae_val, 3), "°C\n")

#scatter plot
ggplot(gage_matched, aes(x = temperature, y = temp_pred)) +
  geom_abline(lty = 2) +
  geom_point(alpha = 0.7) +
  theme_bw() +
  labs(
    x = "USGS Gage Temp (°C)",
    y = "Interpolated Temp (°C)",
    title = paste("Gage vs. Interp on", selected_date)
  )
```

#20 years of data plot
```{r}

#  Define the station_key or gage id for river
sturgeon_key <- "09081000"  

sturgeon_gage <- readNWISsite("09081000") %>%
  select(
    id       = site_no,
    lat_gage = dec_lat_va,
    lon_gage = dec_long_va
  )

grid_pts <- df_full_interp %>%
  distinct(station_key, lat.x, lon.x)
sturgeon_key <- grid_pts$station_key[
  which.min(distHaversine(
    c(sturgeon_gage$lon_gage, sturgeon_gage$lat_gage),
    cbind(grid_pts$lon.x,   grid_pts$lat.x)
  ))
]

# Create a 20-year window 
start_year <- 2003
end_year   <- 2022  

#Pull full daily interpolated temps for Sturgeon River
sturgeon_full <- df_full_interp %>%
  filter(station_key == sturgeon_key,
         year(date) >= start_year,
         year(date) <= end_year)

#Select only the odd days of interest (day of month 7, 11, or 26)
sturgeon_odd <- sturgeon_full %>%
  filter(day(date) %in% c(7, 11, 26)) %>%
  mutate(
    dom = day(date),    # day of month
    yr  = year(date)
  ) %>%
  select(station_key, date, yr, dom, temp_pred)

#Plot 20-year time series for each odd day
ggplot(sturgeon_odd, aes(x = yr, y = temp_pred, group = 1)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ dom, scales = "free_y",
             labeller = labeller(dom = c(`7` = "7th", `11` = "11th", `26` = "26th"))) +
  theme_bw() +
  labs(
    x = "Year",
    y = "Interpolated Temp (°C)",
    title = "Roaring Fork River: Temperature on 7th, 11th, and 26th of Each Month (2003–2022)",
    subtitle = "One point per year per day‑of‑month"
  )

#Create a summary table: min, mean, max across years for each odd day
sturgeon_summary <- sturgeon_odd %>%
  group_by(dom) %>%
  summarize(
    n_years   = n_distinct(yr),   
    min_temp  = min(temp_pred),
    mean_temp = mean(temp_pred),
    max_temp  = max(temp_pred),
    sd_temp   = sd(temp_pred)
  )

print(sturgeon_summary)

```
#quantile plot for single river(gage vs interpolation)
```{r}

gage_key <- "09081000" 
#Use only one gage from river
sturgeon_gage <- readNWISsite("09081000") %>%
  select(lat_gage = dec_lat_va, lon_gage = dec_long_va)
grid_pts <- df_full_interp %>% distinct(station_key, lat.x, lon.x)
sturgeon_key <- grid_pts$station_key[
  which.min(distHaversine(
    c(sturgeon_gage$lon_gage, sturgeon_gage$lat_gage),
    cbind(grid_pts$lon.x, grid_pts$lat.x)
  ))
]

# Build daily residuals
compare_daily <- df_full_interp %>%
  filter(station_key == sturgeon_key) %>%
  select(date, temp_pred_interp = temp.mod) %>%   # pull & rename your interpolated series
  inner_join(
    merged_data_weatherSR %>%
      select(date, temp_true = temp.mod) %>%     # pull the true TempEST2 values
      filter(!is.na(temp_true)),
    by = "date"
  ) %>%
  mutate(
    doy   = yday(date),
    resid = temp_pred_interp - temp_true
  )

# Summarize residual quantiles by day‐of‐year
resid_summary <- compare_daily %>%
  group_by(doy) %>%
  summarize(
    q10   = quantile(resid, 0.10, na.rm=TRUE),
    q25   = quantile(resid, 0.25, na.rm=TRUE),
    med   = median(resid,        na.rm=TRUE),
    q75   = quantile(resid, 0.75, na.rm=TRUE),
    q90   = quantile(resid, 0.90, na.rm=TRUE),
    .groups = "drop"
  )

# Plot bias ribbon
ggplot(resid_summary, aes(x = doy)) +
  geom_ribbon(aes(ymin = q10, ymax = q90), fill = "gray80", alpha = 0.5) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "gray60", alpha = 0.7) +
  geom_line(aes(y = med), color = "firebrick", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_continuous(breaks = c(1, 91, 182, 274, 365),
                     labels = c("Jan","Apr","Jul","Oct","Dec")) +
  theme_bw() +
  labs(
    x     = "Day of Year",
    y     = "Interpolation Bias (°C)",
    title    = paste0("Roaring Fork River (Gage ", gage_key, "): Seasonal Bias of Linear Interpolation"),
    subtitle = "Median residual (solid) with 25–75% (dark) and 10–90% (light) bands"
  )
overall_bias <- resid_summary %>%
  summarize(
    mean_med     = mean(med),
    sd_med       = sd(med),
    mean_abs_med = mean(abs(med))
  )

# Proportion of the year we over‐predict vs under‐predict
prop_sign <- resid_summary %>%
  summarize(
    prop_warm = mean(med > 0),   # fraction of days where resid > 0 (too warm)
    prop_cool = mean(med < 0)    # fraction where resid < 0 (too cool)
  )

# Seasonal breakdown by quarter
seasonal_bias <- resid_summary %>%
  mutate(
    quarter = case_when(
      doy <=  90 ~ "Winter (Q1)",
      doy <= 181 ~ "Spring (Q2)",
      doy <= 273 ~ "Summer (Q3)",
      TRUE       ~ "Fall   (Q4)"
    )
  ) %>%
  group_by(quarter) %>%
  summarize(
    mean_med     = mean(med),
    sd_med       = sd(med),
    mean_abs_med = mean(abs(med)),
    .groups = "drop"
  )

#print metrics
overall_bias
prop_sign
seasonal_bias
range(compare_daily$date)

```
#function to get the gage ids from the data frame
```{r}
# function to extract unique IDs
get_gage_ids <- function(df) {
  if(!"id" %in% names(df)) stop("No `id` column in this data frame")
  unique(df$id)
}

# Apply it to each of your four gage tables
ids_rf  <- get_gage_ids(gage_dataSR)  # Roaring Fork
ids_dr  <- get_gage_ids(gage_data2)   # Delaware
ids_sr  <- get_gage_ids(gage_data3)   # Sturgeon
ids_pr  <- get_gage_ids(gage_data4)   # Potomac

# Combine into a named list 
all_ids <- list(
  RoaringFork = ids_rf,
  Delaware    = ids_dr,
  Sturgeon    = ids_sr,
  Potomac     = ids_pr
)

print(all_ids)

```
#tempest2 quantile plot(no interpolation)
```{r}
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)

# 1. Define the four station_keys you care about
my_keys <- c(
  "47.0322_-88.4918",  
  "40.5648_-75.1734",  
  "39.0941_-106.6550", 
  "39.5207_-78.5689"
)


# Compute quantiles by river × day‐of‐year
quantiles_by_river <- df_keyed_all %>%
  filter(station_key %in% my_keys) %>%        # keep only our four rivers
  mutate(doy = yday(date)) %>%                # compute Day‑of‑Year
  group_by(station_key, doy) %>%              
  summarize(
    q10 = quantile(temp.mod, 0.10, na.rm = TRUE),
    q50 = quantile(temp.mod, 0.50, na.rm = TRUE),
    q90 = quantile(temp.mod, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

# Draw the plot
ggplot(quantiles_by_river, aes(x = doy, color = station_key, fill = station_key)) +
  # 10–90% ribbon
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.2, color = NA) +
  # median line
  geom_line(aes(y = q50), size = 0.8) +
  # formatting
  scale_x_continuous(breaks = c(1,91,182,274,365),
                     labels = c("Jan","Apr","Jul","Oct","Dec")) +
  theme_bw() +
  labs(
    x     = "Day of Year",
    y     = "TempEST2 Model Prediction (°C)",
    color = "River (station_key)",
    fill  = "River (station_key)",
    title = "Seasonal Quantiles of TempEST2 Model Outputs",
    subtitle = "Median (line) with 10–90% range (ribbon) for four rivers"
  )
# gage list
gage_list <- list(
  RoaringFork = gage_dataSR,
  Delaware    = gage_data2,
  Sturgeon    = gage_data3,
  Potomac     = gage_data4
)

# One function to process each river
process_gage <- function(df, river_name) {
  # Parse date
  df <- df %>% mutate(date = as.Date(date))

  # Unique gauge IDs & fetch coords
  ids <- unique(df$id)
  meta <- readNWISsite(ids) %>% 
    select(id = site_no, lat_gage = dec_lat_va, lon_gage = dec_long_va)

  df <- df %>% left_join(meta, by = "id")

  # Build grid of model points
  grid_pts <- df_keyed_all %>% distinct(station_key, lat.x, lon.x)

  # Map each gauge ID to its nearest station_key
  gauge_keys <- df %>%
    distinct(id, lat_gage, lon_gage) %>%
    rowwise() %>%
    mutate(
      station_key = grid_pts$station_key[
        which.min(distHaversine(
          c(lon_gage, lat_gage),
          cbind(grid_pts$lon.x, grid_pts$lat.x)
        ))
      ]
    ) %>%
    ungroup()

  # Pull model predictions by joining on station_key & date
  df_pred <- gauge_keys %>%
    select(id, station_key) %>%
    left_join(
      df_keyed_all %>% select(station_key, date, temp_pred = temp.mod),
      by = "station_key"
    )

  # Now join with the true gauge readings on id & date
  df_comp <- df %>%
    inner_join(df_pred, by = c("id", "date")) %>%
    filter(!is.na(temperature), !is.na(temp_pred))

  # Compute error metrics
  df_comp %>%
    group_by(river = river_name, id) %>%
    summarize(
      RMSE = rmse(temperature, temp_pred),
      MAE  = mae (temperature, temp_pred),
      .groups = "drop"
    )
}

# Apply to all rivers
gage_metrics <- imap_dfr(gage_list, process_gage)

print(gage_metrics)


```

```{r}
filtered_data <- merged_data %>%
  filter(date == as.Date("2023-07-31"))
ggplot(filtered_data, aes(x = lon.x, y = lat.x, color = temp.mod, size = mean_cloud_cover)) +
  geom_point() +
  scale_color_viridis_c(option = "magma", name = "Temperature (C)") +
  scale_size_continuous(range = c(2, 10), name = "Cloud Cover") +
  theme_minimal() +
  labs(
    title = "Temperature vs Cloud Cover on July 2, 2020",
    x = "Longitude",
    y = "Latitude",
    color = "Temperature (°C)",
    size = "Mean Cloud Cover (%)"
  )
```
```{r}
filtered_data2 <- merged_data %>%
  filter(date == as.Date("2023-07-31"))
ggplot(filtered_data2, aes(x = lon.x, y = lat.x, color = temp.mod, size = mean_cloud_cover)) +
  geom_point() +
  scale_color_viridis_c(option = "magma", name = "Temperature (C)") +
  scale_size_continuous(range = c(2, 10), name = "Cloud Cover") +
  theme_minimal() +
  labs(
    title = "Temperature vs Cloud Cover on July 15, 2020",
    x = "Longitude",
    y = "Latitude",
    color = "Temperature (°C)",
    size = "Mean Cloud Cover (%)"
  )
```
```{r}
dates <- as.Date(c("2023-07-28", "2023-07-29", "2023-07-30", "2023-07-31"))

generate_plot <- function(plot_date) {
  filtered_data <- merged_dataSR%>%
    filter(date == plot_date)
  
  ggplot(filtered_data, aes(x = lon.x, y = lat.x, color = temp.mod, size = mean_cloud_cover)) +
    geom_point() +
    scale_color_viridis_c(option = "magma", name = "Temperature (C)") +
    scale_size_continuous(range = c(2, 10), name = "Cloud Cover") +
    theme_minimal() +
    labs(
      title = paste("Temperature vs Cloud Cover on", plot_date),
      x = "Longitude",
      y = "Latitude",
      color = "Temperature (°C)",
      size = "Mean Cloud Cover (%)"
    )
}

# Generate the four plots
plot_0702 <- generate_plot(as.Date("2023-07-02"))
plot_0715 <- generate_plot(as.Date("2023-07-15"))
plot_0716 <- generate_plot(as.Date("2023-07-16"))
plot_0717 <- generate_plot(as.Date("2023-07-17"))

combined_plot <- (plot_0702 | plot_0715) / (plot_0716 | plot_0717)

combined_plot
```
#single day
```{r}
selected_date <- as.Date("2022-07-31")
daily_data <- pred_filtered %>%
  filter(date == selected_date)

daily_data_summary <- daily_data %>%
  group_by(id, lon, lat) %>%
  summarize(
    daily_meanT = mean(temp.mod, na.rm = TRUE),
    daily_range = max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(daily_data_summary, aes(lon, lat, size = daily_range, color = daily_meanT)) +
  geom_point() +
  scale_size_continuous(range = c(2, 10)) +
  scale_color_viridis_c(option = "magma") +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Mean Daily Temperature (C)",
    size = "Temperature Range (C)",
    title = paste("Daily Temperature on", format(selected_date, "%Y-%m-%d"))
  )
```

```{r message=FALSE}
pred_filtered <- predSR %>%
  group_by(id) %>%
  filter(!all(is.na(temp.mod))) %>%
  ungroup()
daily_data <- pred_filtered %>%
# Change to a specfic month
  filter(format(date, "%Y-%m") == "2022-07")

# Group by day, longitude, and latitude, then calculate the daily mean temperature and range
daily_month_data <- daily_data %>%
  group_by(id, date, lon, lat) %>%
  summarize(daily_meanT = mean(temp.mod, na.rm = TRUE),
            daily_range = max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE)) %>%
  ungroup()

ggplot(daily_month_data, aes(lon, lat, size = daily_range, color = daily_meanT)) +
  geom_point() +
  scale_size_continuous(range = c(2, 10)) +
  scale_color_viridis_c(option = "magma") +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Mean Daily Temperature (C)",
    size = "Temperature Range (C)",
    title = "Daily Temperature in May 2021 vs Distance Downstream"
  ) +
  facet_wrap(~ date, ncol = 4)


```

# interpolated TempEst
```{r}
selected_date <- as.Date("2022-07-30")
daily_data_summary_interpolated <- df_interp %>%
  filter(date == selected_date) %>%
  group_by(station_key, lat.x, lon.x) %>%
  summarize(
    daily_meanT = mean(temp.mod, na.rm = TRUE),
    daily_range = max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(daily_data_summary_interpolated, aes(x = lon.x, y = lat.x,
                                            size = daily_range,
                                            color = daily_meanT)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(2, 10), name = "Temperature Range (C)") +
  scale_color_viridis_c(
    option = "magma",
    name   = "Mean Daily Temp (°C)"
  ) +
  coord_quickmap() +
  theme_bw() +
  labs(
    x     = "Longitude",
    y     = "Latitude",
    title = paste("Daily Temperature on", format(selected_date, "%Y-%m-%d"))
  )



```

```{r}
dates_to_check <- as.Date(c("2022-07-01", "2022-07-16", "2022-07-21"))  # example
df_interp %>%
  filter(date %in% dates_to_check) %>%
  group_by(date) %>%
  summarize(
    minT = min(temp.mod, na.rm = TRUE),
    maxT = max(temp.mod, na.rm = TRUE),
    meanT = mean(temp.mod, na.rm = TRUE),
    rangeT = max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE)
  ) %>%
  print(n = Inf)
daily_multi <- df_interp %>%
  filter(date %in% dates_to_check) %>%
  group_by(station_key, lat.x, lon.x, date) %>%
  summarize(
    daily_meanT = mean(temp.mod, na.rm = TRUE),
    daily_range = max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(daily_multi, aes(x = lon.x, y = lat.x, size = daily_range, color = daily_meanT)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ date) +
  scale_size_continuous(range = c(2, 10)) +
    scale_color_viridis_c(
    option = "magma",
    limits = c(20, 30),
    oob = scales::squish,
    name = "Mean Daily Temp (°C)"
  ) +
  coord_quickmap() +
  theme_bw() +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Mean Daily Temperature (C)",
    size = "Temperature Range (C)",
    title = "Daily Temperature Comparison"
  )


```
```{r}
```
monthly_temp_data <- pred_filtered %>% 

mutate(month = format(date, "%Y-%m"), year = format(date, "%Y")) %>% 

group_by(id, month, lon, lat, year) %>% 

summarize(monthly_meanT = mean(temp.mod, na.rm = TRUE), 

monthly_range = max(temp.mod, na.rm = TRUE) - min(temp.mod, na.rm = TRUE)) %>% 

ungroup() 

 

# Filter for the specific year, e.g., 2022 

monthly_temp_data_2022 <- monthly_temp_data %>% 

 

filter(year == "2022") 

 

ggplot(monthly_temp_data_2022, aes(lon, lat, size = monthly_range, color = monthly_meanT)) + 

geom_point() + 

scale_size_continuous(range = c(2, 10)) + 

scale_color_viridis_c(option = "magma") + 

theme_bw() + 

labs( 

x = "Longitude", 

y = "Latitude", 

color = "Mean Monthly Temperature (C)", 

size = "Temperature Range (C)", 

title = "Monthly Temperature in 2022 vs Distance Downstream" 

) + 

facet_wrap(~ month, ncol = 4) 
``` 

# Training a Model

The data requirements are the same, except the data must also have a `temperature`
column for daily mean temperature.  For TempEst 2 development, this was retrieved
from USGS gages.

Using a hypothetical `indat`, then, just run:

```model <- full.schema()(indat)```

Note the empty parentheses; the arguments to `full.schema` are the seasonality
and anomaly model functions to use (defaults provided), and then it returns
another function, which takes training data as an argument, which returns a
model function as above.  This means it is possible to use your own functions
for either of those components.

## Full Demo

For building the PDF, this demo used pre-downloaded data which is too big for
GitHub.  However, the process is shown below, assuming GEE data have been downloaded
to a `GEEData` folder and have IDs matched to USGS gage IDs.

For testing, these data had already been downloaded to `GageTemps.csv`, so that
portion is not run.

```{r}
library(tidyverse)
```


### USGS Data Downloading

Gage data retrieval can fail due to network issues, so it's advisable
to run this in a way that makes it easy to go back and try again.  That is ignored
here.  Running with a timeout is also helpful (`withTimeout` from `R.utils`).

Writing incrementally, instead of collecting a single data frame, helps with
keeping memory use down over many gages and allows the process to be resumed
if it crashes.

```{r}
library(dataRetrieval)
#gage.ids <- c( "01440200","01462000","01434000",
#"01458500","01457500","01438500") 
#gage.ids <- c("01463500", "01440200","01462000","01434000",
#"01458500","01457500","01438500")
gage.ids <- c("09081000", "09085000", "09073300")
gage.file <- "RoaringForkGageData.csv"
start <- "2001-01-01"
end <- "2023-12-31"
var.temp <- "00010"
if (file.exists(gage.file)) {
  file.remove(gage.file)
}
walk(gage.ids, \(gid) {
  
  # Fetch the daily data for the gage
  gage_data <- readNWISdv(gid, var.temp, start, end)
  
  # If data exists for the gage, process and save it
  if (nrow(gage_data) > 0) {
    
    # Rename the columns and format the data
    gage_data_cleaned <- gage_data %>%
      select(id = site_no, date = Date, temperature = X_00010_00003) %>%
      mutate(year = as.integer(format(date, "%Y")),
             day = as.integer(format(date, "%j"))) %>%
      select(id, date, year, day, temperature)
    
    # Write the data to the CSV file (append mode after the first run)
    if (!file.exists(gage.file)) {
      # Write with column names the first time
      write_csv(gage_data_cleaned, gage.file)
    } else {
      # Append without column names after the first run
      write_csv(gage_data_cleaned, gage.file, append = TRUE, col_names = FALSE)
    }
    
  } else {
    cat("No data found for gage ID:", gid, "\n")
  }
  
})
```

```{r}
gage.data <- read_csv(gage.file)
```

```{r}
ggplot(gage.data, aes(date, temperature)) +
  geom_line() +
  theme_bw() +
  labs(
    x="Date",
    y="Daily Mean Temperature (C)"
  )
```


### Load Gage Data

This loads the actual gage data used for analysis, not the demo above.  This has
some technical debt, so it's handled a bit differently to convert it into the
equivalent format above.  There are 6 million observations over 22 years at 1,442
gages, after removing obviously erroneous observations (temperature > 100 or < 0).

```{r}
gage.data <- read_csv("GageTemp.csv", col_types = "cicd") %>%
  extract(time, "day", "([0-9]+)", convert=TRUE) %>%
  mutate(
    date = as.Date(paste0(year, day), format="%Y %j")
  ) %>%
  select(id, date, temperature) %>%
  filter(temperature >= 0, temperature <= 100)
summary(gage.data)
length(unique(gage.data$id))
```

### Load Spatial Data

This has different column structure than the example above, since it was the intial
analysis dataset and variables were retrieved which later turned out to be irrelevant.

The spatial data has 4.6M observations for 1506 points and 22 years.

```{r eval=FALSE}
gee_data_dir <- "data/GreenRiverData"

# List all files in the directory to see what is available
files <- list.files(gee_data_dir, full.names = TRUE)
print(files)

# If there are files in the directory, proceed with reading and processing
if (length(files) > 0) {
  # Print the first few lines of a sample file to verify its contents
  sample_file <- files[1]
  sample_data <- read_csv(sample_file, col_types = "cDddddddddddddddddddd", n_max = 10)
  print(colnames(sample_data))  # Check the column names in the sample file

  # Read and combine all GEE data files into a single data frame
  spat.data <- map_dfr(
    files,
    ~ read_csv(., col_types = "cDddddddddddddddddddd")
  )

  # Print the combined data frame column names to verify
  print(colnames(spat.data))

  # Adjust the script according to the actual column names found in the data
  spat.data <- spat.data %>%
    select(id, date, lat, lon, elevation, lst, humidity, water, shrubland, grassland, barren) %>%
    mutate(
      year = as.integer(format(date, "%Y")),
      day = as.integer(format(date, "%j"))
    ) %>%
    filter(lst > -273, humidity >= 0) %>%  # Remove missing data
    drop_na()

  # Save the processed spatial data for future use
  write_csv(spat.data, "SpatialData.csv")
} else {
  print("No files found in the directory.")
}
```
```{r}
spat.data <- read_csv("SpatialData.csv",
                      col_types = "cDddddddddddii") %>%
  # There are two redundant gages where an old gage was replaced.
  # Kriging does not work with non-unique points.  Easy solution: remove the
  # older gage.
  filter(
    !(id %in% c("420853121505500", "421209121463000"))
  )
```
```{r}
summary(spat.data)
length(unique(spat.data$id))
```


### Combine Data

For training the model, do *not* remove NAs from the combined dataset initially.
The model uses a variety of monthly and long-term means in the spatial data, so
spatial coverage that is not matched to gage coverage remains valuable, and gage
coverage not matched to spatial coverage is useful for fitting seasonal conditions.

The full dataset contains ~8.4 million observations, but only ~2.2 million of those
have fully overlapping coverage.  Many gages do not have continuous coverage for
the full period of interest, and it is nearly guaranteed that some days and points
will not have satellite coverage (cloudy).

```{r eval=FALSE}
data <- full_join(spat.data, gage.data,
                  by=c("id", "date"))
write_csv(data, "AllData.csv")
```
```{r}
data <- read_csv("AllData.csv", col_types = "cDdddddddddiid")
```

### Building a Model

Usually, the objective would be to build a predictive model, in which case we
can leave the arguments to `full.schema` empty (defaults). In this case, we
add the `rtn.model=TRUE` argument, which returns a list of the model components
instead so those can be inspected.  This version does not support prediction.

The default model components are several kriging models, so we can inspect
meaningful coefficients directly.  For example, this could be used to retrieve
the trend in mean annual temperature with respect to river width (approximated by
water abundance), all else being equal.

```{r warning=FALSE, message=FALSE}
model <- full.schema(rtn.model=TRUE)(data)
save(model, file="krigs.rda")
```

To build one that supports prediction:

```{r eval=FALSE}
model <- full.schema()(data)
save(model, file="model.rda")
```

### Inspecting Model Components

Using `rtn.model=TRUE`, each entry in the list is a kriging model.  Some items
of interest are shown below, followed by code to summarize model parameters.

```{r}
load("krigs.rda")
```


```{r}
model$Intercept$summary
summary(model$Intercept)$fixedEffectsTable  # intercept, lon, lat, then fixed terms
colnames(model$Intercept$Z)
```


```{r}
modstats <- map_dfr(model,
                    ~as_tibble_row(.x$summary),
                    .id = "Component")
write_csv(modstats, "SpatialModel.csv")
select(modstats, Component, aRange, sigma2, lambda, tau)
summary(select(modstats, aRange, sigma2, lambda, tau))
```

```{r}
modcoefs <- map_dfr(model,
                    \(mod) {
                      as_tibble(summary(mod)$fixedEffectsTable) %>%
                        mutate(Variable = c("Intercept", "lon", "lat",
                                            colnames(mod$Z))) %>%
                        relocate(Variable)
                    },
                    .id = "Component")
write_csv(modcoefs, "FixedModel.csv")
modcoefs
```

#### Plot Spatial Signal

`fields` includes functions to directly plot a grid of the spatial signal in
a spatial-statistical model.  In effect, this displays spatial anomalies, where
there is a spatial autocorrelation in the component of the model that is not
explained by the fixed regression - in other words, where the sites in a region
tend to behave weirdly (distinct from random noise at an individual site).
This suggests where there may be non-modeled phenomena impacting stream temperature.

The plots are a bit difficult to see embedded in the notebook, so to summarize:

- Intercept (mean temperature) basically has a north-south gradient with scattered
short-distance anomalies.
- Amplitude (annual variation) increases generally to the northeast, but has
several identifiable anomaly regions: northern Arkansas, northern Florida,
northern Arizona, eastern California, and northwestern Oregon.  Several of these
correspond to cross-validation error hot spots.
- The autumn/winter coefficient weakly increases to the west, but the range is
dominated by spatial signal.  There are several high points in the west and low
points in the east.
- The spring/summer coefficient generally increases to the northwest.  Anomalies
are mostly short-ranged, but there is a bit of a general high point in the northern
Rockies/Cascades region.
- The peak winter date has very little discernible gradient; it is dominated by
longer-ranged spatial anomalies.  These are notably present along the West Coast
mountains (early) and near the Great Lakes area (late), along with many weaker
anomalies.
- LST sensitivity increases to the north and a little to the west.  There are
no visible anomalies.
- Humidity sensitivity increases to the west and a little to the north.  There
are no visible anomalies.

```{r}
usa <- map_data("state")

exmod <- model$Humidity

# 25 to 50 N, 125 to 65 W
gList <- list(
  lon = (-125):(-65),
  lat = 25:50
)

# drop.Z: ignore fixed-part covariates
surf <- predictSurface(exmod, gridList = gList, drop.Z = TRUE)
sg <- as_tibble(make.surface.grid(gList)) %>%
  mutate(z = as.vector(surf$z))


ggplot(sg, aes(x=lon, y=lat, fill=z)) +
  geom_tile() +
  geom_polygon(aes(long, lat, group=group, fill=NA), data=usa, color="black",
               fill=NA) +
  theme_bw()
```

Now map it over all of them.

```{r}
spatsig <- map_dfr(model,
                   ~{
                     as_tibble(make.surface.grid(gList)) %>%
                       mutate(z = as.vector(predictSurface(.x,
                                                           gridList = gList,
                                                           drop.Z=TRUE
                                                           )$z))
                   }, .id = "Component")
```

```{r}
spatsig %>%
  drop_na() %>%
  group_by(Component) %>%
  mutate(z = (z - min(z, na.rm=T)) / (max(z, na.rm=T) - min(z, na.rm=T))) %>%
  ggplot(aes(lon, lat, fill=z)) +
  geom_tile() +
  scale_fill_viridis_c() +
  geom_polygon(aes(long, lat, group=group, fill=NA), data=usa, fill=NA, color="black") +
  facet_wrap(~Component) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Normalized Spatial Component"
  )

ggsave("Figures/SpatialSignal.png", width=6.5, height=4, bg="white")
```



#### Plot Fixed Effects

Fixed effect coefficients from before:

```{r}
modcoefs
```

Now extract variable ranges:

```{r}
mc.range <- modcoefs %>%
  filter(Variable != "Intercept") %>%
  rowwise() %>%
  group_modify(~{
    cmp <- first(.x$Component)
    vr <- first(.x$Variable)
    vals <- cbind(model[[cmp]]$Z, model[[cmp]]$x)[,vr]
    y <- model[[cmp]]$y
    mutate(.x,
           YMin = min(y),
           YMax = max(y),
           Min = quantile(vals, 0.05),
           Max = quantile(vals, 0.95))
  }) %>%
  ungroup() %>%
  mutate(
    Effect = estimate * (Max - Min) / (YMax - YMin),
    LogEffect = sign(Effect) * log(abs(Effect) + 1)
  )
```

```{r}
ggplot(mc.range, aes(Variable, Component, fill=Effect)) +
  geom_tile() +
  scale_fill_viridis_c(limits=c(-5, 5), oob=scales::squish) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5)) +
  labs(
    x = "Feature",
    y = "Output",
    fill = "Normalized\nEffect"
  )
ggsave("Figures/LinearEffects.png", width=5, height=3, bg="white")
```

It is important to note that this is *not* showing standalone variable importance,
and key variables may act in combination.  Anywhere that effects are larger than 1,
meaning the effect of the range of the variable exceeds the actual range of the
prediction value, another variable must be counterbalancing it in practice.  For
example, large impacts across humidities suggest that what is happening is probably
a difference between humidities.

### Creating a Recurrent Neural Network Model for TempEst2
```{r}
library(keras)
library(tidyverse)

# Normalize the data
normalized_data <- merged_data_weatherSR %>%
  select(lst, humidity, temperature) %>%
  mutate(across(everything(), ~ (. - mean(.)) / sd(.)))

# Create training and test sets
train_data <- normalized_data %>% filter(date < as.Date("2023-01-01"))
test_data <- normalized_data %>% filter(date >= as.Date("2023-01-01"))library(keras)

# Define the RNN model
model <- keras_model_sequential() %>%
  layer_lstm(units = 50, input_shape = c(10, ncol(train_data) - 1), return_sequences = TRUE) %>%
  layer_lstm(units = 50, return_sequences = FALSE) %>%
  layer_dense(units = 1, activation = "linear")

# Compile the model
model %>% compile(
  loss = "mean_squared_error",
  optimizer = optimizer_adam(),
  metrics = c("mean_absolute_error")
)create_sequences <- function(data, sequence_length) {
  X <- array(NA, dim = c(nrow(data) - sequence_length, sequence_length, ncol(data) - 1))
  y <- array(NA, dim = c(nrow(data) - sequence_length, 1))
  for (i in 1:(nrow(data) - sequence_length)) {
    X[i,,] <- as.matrix(data[i:(i + sequence_length - 1), -ncol(data)])
    y[i] <- data[i + sequence_length, ncol(data)]
  }
  list(X = X, y = y)
}

sequence_length <- 10
train_sequences <- create_sequences(train_data, sequence_length)
test_sequences <- create_sequences(test_data, sequence_length)model %>% fit(
  x = train_sequences$X,
  y = train_sequences$y,
  epochs = 50,
  batch_size = 32,
  validation_split = 0.2
)
predictions <- model %>% predict(test_sequences$X)
```